Testing the Relationship between Phenomenological Control related to Illusion Sensitivity

### 2025-04-28 analysis/1_cleaning.qmd
changes to analysis script

data_ig <- read.csv("https://raw.githubusercontent.com/RealityBending/IllusionGamePhenomenologicalControl/refs/heads/main/data/rawdata_illusion.csv")|> 
  group_by(Illusion_Type) |> 
  mutate(Illusion_Effect = ifelse(sign(Illusion_Strength) > 0, "Incongruent", "Congruent"),
         Task_Difficulty = abs(Illusion_Difference),
         Condition_Illusion = datawizard::categorize(
           Illusion_Strength, split="quantile", n_groups=4,
           labels=c("Congruent - Strong", "Congruent - Mild", "Incongruent - Mild", "Incongruent - Strong")),
         Condition_Difficulty = datawizard::categorize(Task_Difficulty, split="quantile", n_groups=2, labels=c("Hard", "Easy"))) |> 
  ungroup()

@@

data_ppt$pcs_tastescore <- rowMeans(pcs[grepl("taste", names(pcs))])

data_ppt$pcs_postsessionsuggetsion <- sqrt(data_ppt$pcs_urgepress * data_ppt$pcs_memorypress)

#delete with more participants (changes have been made in script)
pcs <- pcs |>
  rename(pcs_amnesia = pcs_rememeber)

# data_ppt$pcs <- rowMeans(pcs[grepl("hands|music|mosquito|tastescore|arm|negative|postsession|amnesia(?!_w)", names(pcs))], na.rm = TRUE)

@@

outliers$Hello <- data_ppt |>
  dplyr::filter(pcs_hello != "Hello")

outliers$Hello <- outliers$Hello$Participant

@@

```{r}
# remove after more pilots
outliers$Balls <- data_ppt |>
  dplyr::filter(pcs_noballs == TRUE) # remove after more pilots

# outliers$balls <- data_ppt |>
#   dplyr::filter(pcs_noballs == "TRUE")

outliers$Balls <- outliers$Balls$Participant

@@

outliers$Press <- data_ppt |>
  dplyr::filter(pcs_press < 5)

outliers$Press <- outliers$Press$Participant
```

@@

```{r}
p_age <- estimate_density(data_ppt$Age) |>
  normalize(select = y) |>
  mutate(y = y * 86) |>  # To match the binwidth
  ggplot(aes(x = x)) +
  geom_histogram(data=dfsub, aes(x = Age, fill=Gender), bins=28) +
  # geom_line(aes(y = y), color = "orange", linewidth=2) +
  geom_vline(xintercept = mean(dfsub$Age), color = "red", linewidth=1.5) +
  # geom_label(data = data.frame(x = mean(df$Age) * 1.15, y = 0.95 * 75), aes(y = y), color = "red", label = paste0("Mean = ", format_value(mean(df$Age)))) +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Age", y = "Number of Participants", color = NULL, subtitle = "Distribution of participants' age") +
  theme_modern(axis.title.space = 10) +
  theme(
    plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
    plot.subtitle = element_text(size = rel(1.2), vjust = 7),
    axis.text.y = element_text(size = rel(1.1)),
    axis.text.x = element_text(size = rel(1.1)),
    axis.title.x = element_blank()
  )
p_age
```

@@

```{r}
p_edu <- data_ppt |>
  mutate(Education = fct_relevel(Education, "Other", "High School", "Bachelor", "Master", "Doctorate")) |>
  ggplot(aes(x = Education)) +
  geom_bar(aes(fill = Education)) +
  scale_y_continuous(expand = c(0, 0), breaks= scales::pretty_breaks()) +
  scale_fill_viridis_d(guide = "none") +
  labs(title = "Education", y = "Number of Participants", subtitle = "Participants per achieved education level") +
  theme_modern(axis.title.space = 15) +
  theme(
    plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
    plot.subtitle = element_text(size = rel(1.2), vjust = 7),
    axis.text.y = element_text(size = rel(1.1)),
    axis.text.x = element_text(size = rel(1.1)),
    axis.title.x = element_blank()
  )
p_edu
```

@@

The final sample includes `r report::report_participants(dfsub, age="Age", sex="Sex", education="Education")`.

>>>>

# data_ig <- read.csv("https://raw.githubusercontent.com/RealityBending/IllusionGamePhenomenologicalControl/refs/heads/main/data/rawdata_illusion.csv")|> 
#   group_by(Illusion_Type) |> 
#   mutate(Illusion_Effect = ifelse(sign(Illusion_Strength) > 0, "Incongruent", "Congruent"),
#          Task_Difficulty = abs(Illusion_Difference),
#          Condition_Illusion = datawizard::categorize(
#            Illusion_Strength, split="quantile", n_groups=4,
#            labels=c("Congruent - Strong", "Congruent - Mild", "Incongruent - Mild", "Incongruent - Strong")),
#          Condition_Difficulty = datawizard::categorize(Task_Difficulty, split="quantile", n_groups=2, labels=c("Hard", "Easy"))) |> 
#   ungroup()

data_ig <- read.csv("C:/Users/asf25/Desktop/studies/IllusionGamePhenomenologicalControl/data/rawdata_illusion.csv")|> 
  group_by(Illusion_Type) |> 
  mutate(Illusion_Effect = ifelse(sign(Illusion_Strength) > 0, "Incongruent", "Congruent"),
         Task_Difficulty = abs(Illusion_Difference),
         Condition_Illusion = datawizard::categorize(
           Illusion_Strength, split="quantile", n_groups=4,
           labels=c("Congruent - Strong", "Congruent - Mild", "Incongruent - Mild", "Incongruent - Strong")),
         Condition_Difficulty = datawizard::categorize(Task_Difficulty, split="quantile", n_groups=2, labels=c("Hard", "Easy"))) |> 
  ungroup()

@@

pcs$pcs_tastescore <- rowMeans(pcs[grepl("taste", names(pcs))])

pcs$pcs_postsessionsuggetsion <- sqrt(data_ppt$pcs_urgepress * data_ppt$pcs_memorypress)

pcs$pcs_postsessionsuggetsion <- sqrt(data_ppt$pcs_urgepress * data_ppt$pcs_memorypress)

@@

# outliers$Hello <- data_ppt |>
#   dplyr::filter(pcs_hello != "Hello")
# 
# outliers$Hello <- outliers$Hello$Participant

@@

```{r}
# remove after more pilots
# outliers$Balls <- data_ppt |>
#   dplyr::filter(pcs_noballs == TRUE) # remove after more pilots
# 
# # outliers$balls <- data_ppt |>
# #   dplyr::filter(pcs_noballs == "TRUE")
# 
# outliers$Balls <- outliers$Balls$Participant
```

@@

# outliers$Press <- data_ppt |>
#   dplyr::filter(pcs_press < 5)
# 
# outliers$Press <- outliers$Press$Participant
```

@@

```{r}
p_age <- estimate_density(data_ppt$Age) |>
  normalize(select = y) |>
  mutate(y = y * 86) |>  # To match the binwidth
  ggplot(aes(x = x)) +
  geom_histogram(data=data_ppt, aes(x = Age, fill=Gender), bins=28) +
  # geom_line(aes(y = y), color = "orange", linewidth=2) +
  geom_vline(xintercept = mean(data_ppt$Age), color = "red", linewidth=1.5) +
  # geom_label(data = data.frame(x = mean(df$Age) * 1.15, y = 0.95 * 75), aes(y = y), color = "red", label = paste0("Mean = ", format_value(mean(df$Age)))) +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Age", y = "Number of Participants", color = NULL, subtitle = "Distribution of participants' age") +
  theme_modern(axis.title.space = 10) +
  theme(
    plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
    plot.subtitle = element_text(size = rel(1.2), vjust = 7),
    axis.text.y = element_text(size = rel(1.1)),
    axis.text.x = element_text(size = rel(1.1)),
    axis.title.x = element_blank()
  )
p_age
```

@@

```{r}
p_edu <- data_ppt |>
  mutate(Education = fct_relevel(Education, "Other", "Elementary school", "High school", "Bachelor", "Master", "Doctorate")) |>
  ggplot(aes(x = Education)) +
  geom_bar(aes(fill = Education)) +
  scale_y_continuous(expand = c(0, 0), breaks= scales::pretty_breaks()) +
  scale_fill_viridis_d(guide = "none") +
  labs(title = "Education", y = "Number of Participants", subtitle = "Participants per achieved education level") +
  theme_modern(axis.title.space = 15) +
  theme(
    plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0),
    plot.subtitle = element_text(size = rel(1.2), vjust = 7),
    axis.text.y = element_text(size = rel(1.1)),
    axis.text.x = element_text(size = rel(1.1)),
    axis.title.x = element_blank()
  )
p_edu
```

@@

The final sample includes `r report::report_participants(data_ppt, age="Age", sex="Sex", education="Education")`.

### 2025-04-28 analysis/2_analysis.qmd
changes to analysis script

---
title: "Illusion"
subtitle: "Illusion Task"
date: "`r Sys.Date()`"
editor: source
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
    self-contained: false
    toc: true
execute: 
  cache: true
---

>>>>

---
title: "Illusion Game & PCS"
subtitle: "Illusion Task"
date: "`r Sys.Date()`"
editor: source
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
    self-contained: false
    toc: true
execute: 
  cache: true
---

### 2025-04-28 paper/R1/manuscript.qmd
changes to analysis script

```{r}
#| warning: false
#| label: fig-illusion_framework
#| apa-twocolumn: false  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"
#| fig-cap: "The parametric framework for visual illusions (Makowski et al., 2021) applied to the Müller-Lyer illusion. Below are examples of stimuli showcasing the manipulation of two parameters, task difficulty and illusion strength."

knitr::include_graphics("img/IG_framework.png")
```

The "Illusion Game" paradigm [@makowski2023] has been recently developed to measure illusion sensitivity to various illusion types through its behavioural impact (on response time and error rate) in a perceptual decision task (where participants have to respond as fast as possible; e.g., "which of the left or right circles is bigger"). 
The stimuli for different classical illusions are created using the *Pyllusion* software [@makowski2021], which allows researchers to modulate the strength of the illusion as a continuous dimension, independently from the difficulty of the perceptual task (see @fig-illusion_framework). 
This paradigm, inspired by psychophysics, lends itself to the computational modelling of illusion sensitivity through its **interference effect —an effect that disrupts an individual's ability to accurately discriminate between perceptual stimuli.**
**This approach aims to bypass some of the metacognitive processes involved in other paradigms, offering a more direct and objective measure of how illusions influence perceptual judgment.**

@@

The task is an adaptation of the one used in @makowski2023 to make it shorter, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal (see @fig-illusionexample).

@@

**In the original Illusion Game, 10 visual illusions were presented in two sets, following a practice trial, and separated by two short questionnaires.**
**Participants completed a total of 1,340 trials, with the experiment lasting approximately 55 minutes.**
**In the current procedure, only three illusions are used, selected based on the original study's findings that these illusions most strongly contribute to illusion sensitivity.**

@@

For each illusion type, two continuous dimensions are orthogonally manipulated namely task difficulty and illusion strength, so that each trial corresponds to a unique combination, **providing an objectively correct answer for each task.**
**The use of these manipulations allows concise, standardised reporting of illusion parameters and ensures our stimuli are fully reproducible [see @makowski2021].**

@@

Illusion Game outliers will be flagged based on their RT distributions, following the same procedure as in [@makowski2023]. 
**If the RT is collapsed to the left (i.e., has \> 1/3 of ultra-fast responses - typically \< 200 ms) in the first set, the entire participant will be discarded (suggesting that they did not properly do the task), but if only the second set is bad, then only the second set will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision).** 
In addition, the removal of individual trials will also be performed [RT \< 200 ms or \> 3 SD; following @theriault2024].
**To mitigate the risk of confounding effects driven by extreme speed or accuracy strategies, participants whose RTs are significantly slower than the group average (RT > 4 SD above the mean, based on @makowski2023) will be excluded from the analysis.**

After removing problematic participants and trials, the outcome measures (PC and VI sensitivity scores) will be computed and the Bayesian correlation (with medium prior on the coefficient, i.e., r-scale parameter set to 1/3) will be computed [using the *BayesFactor* package, @BayesFactor].
This shifted beta prior, as recommended by @morey2018baysefactor, offers a balanced approach to estimating effect sizes, without placing undue weight on larger effect sizes or artificially inflating evidence for the null hypothesis.
Following @lush2022, we expect to collect evidence against (BF10 \<= 1/3) a relationship between PCS and VI sensitivity. 
**Additionally, Bayesian correlations will be computed using the BayesFactor package, employing a medium prior on the coefficient (r-scale parameter set to 1/3) to assess relationships between maladaptive trait facets and illusion sensitivity scores.**
**Based on prior research (Makowski et al., 2023), we expect to find evidence (BF10 ≥ 3) supporting a relationship between the psychoticism facet of the PID-5 and illusion sensitivity.**
Data analysis will be carried out using R, using *tidyverse* [@tidyverse] and *easystats* [@easystats; @parameters; @correlation; @bayestestR; @datawizard]. 
The analysis script and additional information are available at **https://osf.io/da3u6/?view_only=247d4efa1afe456aa07662732946d4e6** [Note this link will be replaced with the GitHub page of the current project upon completion of the review process to ensure continued anonymisation].

>>>>

The "Illusion Game" paradigm [@makowski2023] has been recently developed to measure illusion sensitivity to various illusion types through its behavioural impact (on response time and error rate) in a perceptual decision task (where participants have to respond as fast as possible; e.g., "which of the left or right circles is bigger"). 
The stimuli for different classical illusions are created using the *Pyllusion* software [@makowski2021], which allows researchers to modulate the strength of the illusion as a continuous dimension, independently from the difficulty of the perceptual task. 
This paradigm, inspired by psychophysics, lends itself to the computational modelling of illusion sensitivity through its **interference effect —an effect that disrupts an individual's ability to accurately discriminate between perceptual stimuli.**
**This approach aims to bypass some of the metacognitive processes involved in other paradigms, offering a more direct and objective measure of how illusions influence perceptual judgment.**

@@

The task is an adaptation of the one used in @makowski2023 to make it shorter, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal (see @fig-illusionexample). **In the original Illusion Game, 10 visual illusions were presented in two sets, following a practice trial, and separated by two short questionnaires.**
**Participants completed a total of 1,340 trials, with the experiment lasting approximately 55 minutes.**
**In the current procedure, only three illusions are used, selected based on the original study's findings that these illusions most strongly contribute to illusion sensitivity.**

@@

@@

For each illusion type, two continuous dimensions are orthogonally manipulated namely task difficulty and illusion strength, so that each trial corresponds to a unique combination, **providing an objectively correct answer for each trial.**
**The use of these manipulations allows concise, standardised reporting of illusion parameters and ensures our stimuli are fully reproducible [see @makowski2021].**

@@

Illusion Game outliers will be flagged based on their RT distributions, following the same procedure as in [@makowski2023]. 
**If the RT is collapsed to the left (i.e., has \> 1/3 of ultra-fast responses - typically \< 200 ms) in the first set, the entire participant will be discarded (suggesting that they did not properly do the task), but if only the second set is bad, then only the second set will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision).** 
In addition, the removal of individual trials will also be performed [RT \< 200 ms or \> 3 SD; following @theriault2024].
**To mitigate the risk of confounding effects driven by extreme speed or accuracy strategies, participants whose RTs are significantly slower than the group average [RT > 4 SD above the mean, based on @makowski2023] will be excluded from the analysis.**

After removing problematic participants and trials, the outcome measures (PC and VI sensitivity scores) will be computed and the Bayesian correlation (with medium prior on the coefficient, i.e., r-scale parameter set to 1/3) will be computed [using the *BayesFactor* package, @BayesFactor].
**This shifted beta prior, as recommended by @morey2018baysefactor, offers a balanced approach to estimating effect sizes, without placing undue weight on larger effect sizes or artificially inflating evidence for the null hypothesis.**
Following @lush2022, we expect to collect evidence against (BF10 \<= 1/3) a relationship between PCS and VI sensitivity. 
**Additionally, Bayesian correlations will be computed using the BayesFactor package, employing a medium prior on the coefficient (r-scale parameter set to 1/3) to assess relationships between maladaptive trait facets and illusion sensitivity scores.**
**Based on prior research (Makowski et al., 2023), we expect to find evidence (BF10 ≥ 3) supporting a relationship between the psychoticism facet of the PID-5 and illusion sensitivity.**
Data analysis will be carried out using R, using *tidyverse* [@tidyverse] and *easystats* [@easystats; @parameters; @correlation; @bayestestR; @datawizard]. 
The analysis script and additional information are available at **https://osf.io/da3u6/?view_only=247d4efa1afe456aa07662732946d4e6** [Note this link will be replaced with the GitHub page of the current project upon completion of the review process to ensure continued anonymisation].
